---
title: "2015-02-24-DhruvNair-HW2"
author: "Dhruv Nair"
date: "Tuesday, February 24, 2015"
output: html_document
---
**Assignment**
Recreate the NYT Ebola Graphics in R

####Preprocessing the Data

```{r}
setwd("~/Github/data-viz/ebola")
Microbe_Data_Raw <- read.csv(file="Microbe-scope-bugs.csv", header=TRUE)
```


The data frame contains columns without discriptive names, as well as columns that only contain notes or links. Since none of these columns are contributing to our plot, we can remove them from out data frame. 


```{r warning=FALSE}
##Add  the column name "Microbe.Name" to first column of the data frame
names(Microbe_Data_Raw)[1]<-"Microbe.Name"

##identify the columns in the data frame titled as "notes"
notes <- grep(pattern="note", x=names(Microbe_Data_Raw), ignore.case=TRUE)
names(Microbe_Data_Raw)[notes]

##Remove the selected columns from the data frame
Microbe <- subset(Microbe_Data_Raw, select = -notes)

##Function to drop the  % sign from the case.fatality.rate column and turn it into numeric values
dropPct<-function(df, var){
  dropPct<-strsplit(as.character(df[[var]]), split="%")
  df[[var]]<-as.numeric(dropPct)
  return(df[[var]])
  }

##Implementing the function
Microbe[["case.fatality.rate"]]<-dropPct(Microbe, "case.fatality.rate")

##Converting the average.basic.reproductive.rate to numeric value
Microbe$average.basic.reproductive.rate <- as.numeric(as.character(Microbe$average.basic.reproductive.rate))

```

####Plotting the Data

```{r}
##Import ggplot library
library(ggplot2)
```  

```{r}
##Plot the data 
Germ_Plot <- ggplot(Microbe, aes(x=average.basic.reproductive.rate,y=case.fatality.rate,color=primary.mode.of.transmission,label=Microbe.Name))
```

I couldn't quite figure out how to create a plot with an uneven y-axis (I don't think ggplot supports that functionality). I created two separate plots. One with the y-axis scaled to a log scale, and the second that was just a simple even scale.  Adding the scatterplot layer and scaling the plot to fit the data

```{r}
##Germ_Plot1 plots the smaller case.fatality.rate values on an equally spaced log scale. 
Germ_Plot1<- Germ_Plot + geom_point(shape=19,alpha=1,size=4) + scale_x_continuous(breaks=1:20, limits=c(0,17)) + scale_y_log10(breaks=c(0.01,0.1,1), limits=c(0.01,1)) 

##Germ_Plot2 plots the larger case.fatality.rate values on a continuous y scale ranging from 0-100 
Germ_Plot2 <- Germ_Plot + geom_point(shape=19,alpha=1,size=4) + scale_x_continuous(breaks=1:20, limits=c(0,17)) + scale_y_continuous(breaks=(c(0,seq(10,100,10))), limits=c(0,100))
```

####Add the dividing lines 
```{r}

Germ_Plot1 <- Germ_Plot1 + geom_vline(alpha=1,xintercept=5,colour="red") + geom_vline(alpha=1,xintercept=7,colour="red") + geom_vline(alpha=1,xintercept=12,colour="red")

Germ_Plot2 <- Germ_Plot2 + geom_vline(alpha=1,xintercept=5,colour="red") + geom_vline(alpha=1,xintercept=7,colour="red") + geom_vline(alpha=1,xintercept=12,colour="red")

Germ_Plot1 <- Germ_Plot1 + geom_hline(yintercept=1) + geom_vline(xintercept=1) 
Germ_Plot2 <- Germ_Plot2 + geom_hline(yintercept=1) + geom_vline(xintercept=1) 

##
Germ_Plot2 <- Germ_Plot2 + geom_hline(alpha=1,yintercept=1,linetype="dashed")
Germ_Plot2 <- Germ_Plot2 + geom_hline(alpha=1,yintercept=20,linetype="dashed")
Germ_Plot2 <- Germ_Plot2 + geom_hline(alpha=1,yintercept=50,linetype="dashed")
```
####Add the labels to all the data points and to the axes
```{r}
##
Germ_Plot1 <- Germ_Plot1 + geom_text(hjust=0, vjust=0,size=3)+ xlab("CONTAGIOUSNESS,Average Basic Reproductive Rate") + ylab("DEADLINESS,Case Fatality Rate")
Germ_Plot2 <- Germ_Plot2 + geom_text(hjust=0, vjust=0,size=3) + xlab("CONTAGIOUSNESS,Average Basic Reproductive Rate") + ylab("DEADLINESS,Case Fatality Rate")

```

```{r}
##Expand the plot to make room for annotations, and change the theme to a white background with dashed lines. 
library(grid)
Germ_Plot1 <- Germ_Plot1 + theme(plot.margin = unit(c(3,3,3,3), "lines")) + theme_bw() +  theme(panel.grid=element_line(linetype="dashed")) + theme(legend.position="top")
Germ_Plot2 <- Germ_Plot2 + theme(plot.margin = unit(c(14,14,14,14), "lines")) + theme_bw() +  theme(panel.grid=element_line(linetype="dashed")) + theme(legend.position="top") + labs(title="The Microbe-scope")
```

####Adding the contagiousness annotations
```{r}
##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Not Very", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=1.25,ymax=1.25,xmin=0.25,xmax=0.25)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Not Very", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=103,ymax=103,xmin=0.25,xmax=0.25)

##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Quite Contagious", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=1.25,ymax=1.25,xmin=3,xmax=3)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Quite Contagious", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=103,ymax=103,xmin=3,xmax=3)

##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Very", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=1.25,ymax=1.25,xmin=6,xmax=6)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Very", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=103,ymax=103,xmin=6,xmax=6)

##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Highly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=1.25,ymax=1.25,xmin=9.5,xmax=9.5)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Highly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=103,ymax=103,xmin=9.5,xmax=9.5)

##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Vaccinate Now!", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=1.25,ymax=1.25,xmin=14.5,xmax=14.5)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Vaccinate Now!", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=103,ymax=103,xmin=14.5,xmax=14.5)

```
####Add the Deadliness annotations
```{r}
##
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Extremely Deadly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=80,ymax=80,xmin=14.5,xmax=17)

##
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Deadly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=35,ymax=35,xmin=14.5,xmax=17)

##
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Quite Deadly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=15,ymax=15,xmin=14.5,xmax=17)

##
Germ_Plot1<- Germ_Plot1 + annotation_custom(grob=textGrob(label = "Not Too Deadly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=0.001,ymax=0.01,xmin=14.5,xmax=17)
Germ_Plot2<- Germ_Plot2 + annotation_custom(grob=textGrob(label = "Not Too Deadly", hjust = 0, gp = gpar(fontsize=8,cex = 1.5)), ymin=0.01,ymax=0.01,xmin=14.5,xmax=17)
```


```{r fig.width=15, fig.height=12, warning=FALSE}
Germ_Plot1
Germ_Plot2

```

