---
title: "2015-02-05-XuetingWu-Assignment1"
author: "Xueting Wu"
date: "February 04, 2015"
output: html_document
---

**1.Go to this link to retrieve general metadata for Local Government Areas (LGAs) of Nigeria. LGAs are akin to counties in the U.S.**
```{r}
setwd("/Users/xuetingwu/data-viz/r")
#By forking the upstream repository, the general metadata for LGAs is already on the local drive. So just assign a name "data1" to the .csv
data1 <- read.csv("/Users/xuetingwu/data-viz/r/lgas.csv")
```
Explore the data a little bit to get familiar with it:

(1)How is the data organized? 
```{r}
head(data1)  # first 6 observations of the dataset.
str(data1)  # a more programming-oriented sense including the class of each element
```
(2)How big is the data?
```{r}
dim(data1) # the dimension of the dataset
```
(3)How complete is the data?
```{r}
sum(is.na(data1))  #count the number of NAs in the LGAs dataset
```


**2. Next, download the full NMIS dataset, and grab the [large] .csv file called “Health_Mopup_and_Baseline_NMIS_Facility.csv”. This contains every health facility in Nigeria.**

```{r}
data2 <- read.csv("/Users/xuetingwu/data-viz/r/nmis_dataset/Health_Mopup_and_Baseline_NMIS_Facility.csv") # import the data
# get a general idea about the data
head(data2)
str(data2)
dim(data2)
```

**3. Produce a new data.frame with the following properties:**

(1)From the NMIS Health Facilities Inventory, select all facilities located in the Southern zones of Nigeria.
```{r}
table(data1$zone)
southz <- subset(data1,zone %in% c("South-South","Southeast","Southwest")) # select the southern zone from LGAs metadata
dim(southz)
122+95+137 # check the subsetting process
fsouthz <- subset(data2, unique_lga %in% southz$unique_lga) # select facilities located in the Southern zones of Nigeria according to "unique_lga" in dataset "southz"
```

(2)Incorporate LGA metadata into this new dataframe.
```{r}
m <- merge(southz, fsouthz, by="unique_lga") 
dim(fsouthz)-dim(m)  # check the merging process and the result is as expected
```
check the operation of building the new data.frame using another approach:
```{r}
data3 <- merge(data1, data2, by="unique_lga")
dim(data2)-dim(data3)  # check the merging process. Considering the number of rows was reduced by 948, there are 948 observations in data2 whose unique_lga don't match any of the unique_lga in data1 possibily due to typing errors.
m2 <- subset(data3,zone =="South-South"| zone=="Southeast"| zone=="Southwest")  # select facilities in the southern zone from the new dataset
dim(m)-dim(m2)  # compare the two data.frames which are supposed to be equal
```

**4. Calculate the total number of full time nurses and doctors for all health facilities, by state. Compute per capita and per area values, as well.**

(1)Sort the resulting dataset by state population, in descending order.
```{r}
library(plyr)
# to get rid of NAs without having to remove the whole observation
data3$num_nurses_fulltime[is.na(data3$num_nurses_fulltime)]=0
data3$num_nursemidwives_fulltime[is.na(data3$num_nursemidwives_fulltime)]=0
data3$num_doctors_fulltime[is.na(data3$num_doctors_fulltime)]=0
data3$fultotal <- data3$num_nurses_fulltime + data3$num_nursemidwives_fulltime+data3$num_doctors_fulltime  #nursemidwives are also counted as fulltime nurses
which(colnames(data3)=="fultotal")
Fulltime_Total<-aggregate(data3[43],list(data3$state),sum)
which(colnames(data1)=="pop_2006")
State_Pop <- aggregate(data1[3], list(data1$state), sum)
which(colnames(data1)=="area_sq_km")
State_Area_Sq_Km <- aggregate(data1[7], list(data1$state), sum)
Newdata1 <- merge(Fulltime_Total, State_Pop, by= "Group.1" )
Newdata2 <- merge(Newdata1, State_Area_Sq_Km, by= "Group.1" )
names(Newdata2)[1] <- "state"
Newdata3 <- transform(Newdata2, new = fultotal / pop_2006)
Newdata2$per_capita <- Newdata2$fultotal/Newdata2$pop_2006
Newdata2$per_area_unit <- Newdata2$fultotal/Newdata2$area_sq_km
Newdata_Sorted <- arrange(Newdata2, desc(pop_2006))
```
  (2)Show the results!
```{r}
print(Newdata_Sorted)
library(ggplot2)
ggplot(Newdata_Sorted, aes(x=state, y=per_capita))+geom_point()
ggplot(Newdata_Sorted, aes(x=state, y=per_area_unit))+geom_point()
ggplot(Newdata_Sorted, aes(x=per_capita, y=per_area_unit, colours=state))+geom_point() 
```